GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations

### 2022-12-05 writeUps/paper.tex
update writeup

\begin{prop}\label{prop:reg1}
  Under Assumptions \ref{ass:sm}, \ref{ass:rand}, \ref{ass:vps}, and \ref{ass:ci}, if %principal scores are known exactly,
  $\ri$ is defined as in \eqref{eq:ri} and the model
\begin{equation}\label{eq:regression0}
  \EE[Y_i|Z_i,\ri]=\beta_0+\beta_1\ri+\beta_2Z_i+\beta_3Z_i\ri
\end{equation}
is fit with OLS, yielding coefficient estimates $\hat{\beta}_k$, $k=0,\dots,3$, then
\begin{equation}
  \begin{split}
    \hat{\eff0}&\equiv \hat{\beta}_2\\
    \hat{\eff1}&\equiv \hat{\beta}_3-\hat{\beta}_2
  \end{split}
\end{equation}
are M-estimators, with $\hat{\eff0}\rightarrow_p\eff0$ and $\hat{\eff1}\rightarrow_p\eff1$ as $n\rightarrow\infty$
\end{prop}

@@

The mixture estimator then fit two outcome models.
For the treatment group, a normal regression:
\begin{equation}\label{eq:mix1-sim}
  f_{Y|Z=1}(y_i)=\phi\left(\frac{y_i-[\beta_0+\beta_1S_i+\beta_2 +\bm{\gamma}_1^T\bm{x}_{i,1:2}]}{\sigma_T}\right)
\end{equation}
and for the control group, mixture model:
\begin{equation}\label{eq:mix-sim}
  f_{Y|Z=0}(y_i)=\ppf{\bm{\alpha}}{\bm{x}_{i,1:2}}\phi\left(\frac{y_i-[\beta_0+\beta_1+\bm{\gamma}_1^T\bm{x}_{i,1:2}]}{\sigma_C}\right)}+(1-\ppf{\bm{\alpha}}{\bm{x}_{i,1:2}})\phi\left(\frac{y_i-[\beta_0+\bm{\gamma}_1^T\bm{x}_{i,1:2}]}{\sigma_C}\right)}
\end{equation}
Note that the coefficients $\bm{\gamma}_1$ are the same across treatment groups and principal strata, while the residual variance differs across treatment groups, but not across principal strata.
The models \eqref{eq:psModsim}, \eqref{eq:mix1-sim}, and \eqref{eq:mix-sim} were fit simultaneously using Bayesian Markov Chain Monte Carlo with Stan \citep{rstan}. We interpreted the posterior mean as a point estimate and the posterior standard deviation as a standard error.

@@

There is good reason to hope that extensions to GEEPERs may be straightforward.
First, GEEPERs may be extended to situations in which there is two-way noncompliance, and only a weaker form of monotonicity (Assumption \ref{ass:sm}) holds.
For instance, in truncation by death problems \citep{e.g.][]{zhangRubin,ding2011}, $S_i=1$ if participant $i$ survives (or, more generally, if the outcome $Y_i$ is measured) and 0 otherwise, and interest is is in the principal effect for the ``always survive'' principal stratum in which $\sti=S_{ci}=1$.
If, say, $\sti\ge S_{Ci}$, then every subject in the control condition with $S_i=1$ is in the always survive stratum, while those subjects in the treatment condition with $S_i=1$ are a mixture of the always survive and $\sti=1;\;S_{Ci}=0$ stratum.
The scenario of truncation by death with (weak) monotonicity is broadly similar to the strong monotonicity, one-way noncompliance situation that this paper discussed.
On the other hand, when no monotonicity assumption holds, both $\st$ and $S_C$ will have to be imputed for every subject; further research is necessary to determine the appropriate way to do so.

@@

\subsection{Proof for Proposition \ref{prop:reg0}}
Carrying out those replacements in \eqref{eq:estEq0}, and rearranging\footnote{Technically, replacing $\tilde{\Psi}_i$ with $\Psi_i=\begin{psmallmatrix} 1 & 0&1&0\\ 0&1&0&1\\0&0&1&0\\0&0&0&1\end{psmallmatrix}\tilde{\Psi}_i$.} gives an equivalent set of estimating equations:
\begin{equation}\label{eq:estEq1}
\sum_{i=1}^{n_C}\Psi_i=\sum_{i=1}^{n_C}\begin{pmatrix}
    Y_i-\muc0-\ri(\muc1-\muc0)-Z_i(\mut0-\muc0)-Z_i\ri(\mut1-\mut0-\muc1+\muc0)\\
    \ri Y_i-\ri\muc0-\ri^2(\muc1-\muc0)-Z_i\ri(\mut0-\muc0)-Z_i\ri^2(\mut0-\mut1-\muc0+\muc1)\\
    Z_iY_i-Z_i\mut0-Z_i\ri (\mut1-\mut0)\\
    Z_i\ri Y_i -Z_i\ri\mut0-Z_i\ri^2(\mut1-\mut0)

\end{pmatrix}=\begin{pmatrix} 0\\0\\0\\0\end{pmatrix}
\end{equation}

\subsection{Proof for Proposition \ref{prop:interaction}}
First, note that $\EE[S]=\EE\EE[S|\bx]=\EE[\pp]$ and
$$\EE[\bx S]=\EE[\bx S|Z=1]=\EE[\bx S|Z=0]=\EE[\bx \EE[\bx\EE[S|\bx]|Z=0]]=\EE[\bx\pp|Z=0]=\EE[\bx\pp]$$
implying that
\begin{equation*}
  \begin{split}
    \EE[Y|\bx,Z=0]&=\beta_0+\bm{\beta_1}\bxy+\beta_2\pp+\bm{\beta_5}'\bxy\pp\\
    \EE[Y|\bx,S,Z=1]&=\beta_0+\beta_3+(\bm{\beta_1}+\bm{\beta_6})\bxy+(\beta_3+\beta_4)S+(\bm{\beta_5}+\bm{\beta_7})'\bxy S
  \end{split}
\end{equation*}
Therefore,
\begin{align*}
  \EE[Y]&=\EE[Y|Z=0]+\EE[Z]\left\{\EE[Y|Z=1]-\EE[Y|Z=0]\right\}\\
  &=\beta_0+\bm{\beta_1}\EE[\bxy]+\beta_2\EE[\pp]+\bm{\beta_5}'\EE[\bxy\pp]+\EE[Z]\left\{\beta_3+\bm{\beta_4}\EE[S]+\bm{\beta_6}'\EE[\bxy]+\bm{\beta_7}'\EE[\bxy S]\right\}\\
  &=\beta_0+\bm{\beta_1}\EE[\bxy]+\beta_2\EE[R]+\bm{\beta_5}'\EE[\bxy R]+\EE[Z]\left\{\beta_3+\bm{\beta_4}\EE[R]+\bm{\beta_6}'\EE[\bxy]+\bm{\beta_7}'\EE[\bxy R]\right\}\\
\end{align*}
Analogous reasoning leads to expressions for $\EE[RY]$, $\EE[\bxy Y]$, $\EE[ZY]$,  $\EE[ZRY]$, $\EE[\bxy YZ]$, and $\EE[\bxy RZY]$.
These, in turn, give rise to estimating equations
\begin{equation*}
  \displaystyle\sum_i \begin{pmatrix}
    Y_i-\beta_0-\bm{\beta_1}\bxy_i-\beta_2R_i-\bm{\beta_5}'\bxy_iR_i-Z_i\left\{\beta_3-\bm{\beta_4}R_i-\bm{\beta_6}'\bxy_i-\bm{\beta_7}'\bxy_i R_i\right\}\\
    R_iY_i-\beta_0R_i-\bm{\beta_1}\bxy_i R_i-\beta_2R_i^2-\bm{\beta_5}'\bxy_iR_i^2-Z_i\left\{\beta_3R_i-\bm{\beta_4}R_i^2-\bm{\beta_6}'\bxy_iR_i-\bm{\beta_7}'\bxy_i R_i^2\right\}\\
    \bxy_i Y_i-\beta_0\bxy_i-\bm{\beta_1}\bxy_i\bx^{Y'}_i-\beta_2R_i\bxy_i-\bm{\beta_5}'\bxy_iR_i-Z_i\left\{\beta_3\bxy_i-\bm{\beta_4}R_i\bxy_i-\bm{\beta_6}'\bxy_i\bx^{Y'}_i-\bm{\beta_7}'\bxy_i\bx^{Y'}_i R_i\right\}\\
    Z_iY_i-Z_i\left\{\beta_0-\bm{\beta_1}\bxy_i-\beta_2R_i-\bm{\beta_5}'\bxy_iR_i-\beta_3-\bm{\beta_4}R_i-\bm{\beta_6}'\bxy_i-\bm{\beta_7}'\bxy_i R_i\right\}\\
    Z_iR_iY_i-Z_iR_i\left\{\beta_0-\bm{\beta_1}\bxy_i-\beta_2R_i-\bm{\beta_5}'\bxy_iR_i-\beta_3-\bm{\beta_4}R_i-\bm{\beta_6}'\bxy_i-\bm{\beta_7}'\bxy_i R_i\right\}\\
    Z_iY_i\bxy_i-Z_i\left\{\beta_0\bxy_i-\bm{\beta_1}\bxy_i\bx^{Y'}_i-\beta_2R_i\bxy_i-\bm{\beta_5}'\bxy_i\bx^{Y'}_iR_i-\beta_3\bxy_i-\bm{\beta_4}R_i\bxy_i-\bm{\beta_6}'\bxy_i\bx^{Y'}_i-\bm{\beta_7}'\bxy_i\bx^{Y'}_i R_i\right\}\\
    R_iZ_iY_i\bxy_i-Z_iR_i\left\{\beta_0\bxy_i-\bm{\beta_1}\bxy_i\bx^{Y'}_i-\beta_2R_i\bxy_i-\bm{\beta_5}'\bxy_i\bx^{Y'}_iR_i-\beta_3\bxy_i-\bm{\beta_4}R_i\bxy_i-\bm{\beta_6}'\bxy_i\bx^{Y'}_i-\bm{\beta_7}'\bxy_i\bx^{Y'}_i R_i\right\}
  \end{pmatrix}=\begin{pmatrix} 0\\0\\0\\0\\0\\0\\0\end{pmatrix}
\end{equation*}
These are the estimating equations for the regression model \eqref{eq:fullInteractionModel}, fit by OLS. Now note that
\begin{equation*}
  \EE[Y_T-Y_C|\st=0]=\beta_0^T-\beta_0^C+(\bm{\beta_1}^{T'}-\bm{\beta_1}^{C'})\EE[\bxy|S=0]=\beta_3+\beta_6'E[\bxy|S=0]
\end{equation*}
and
\begin{equation*}
  \EE[Y_T-Y_C|\st=1]=\beta_0^T+\beta_2^T-(\beta_0^C+\beta_2^C)+(\bm{\beta_1}^{T'}+\bm{\beta_3^{Ti}}-\bm{\beta_1}^{C'}-\bm{\beta_3^{C'}})\EE[\bxy|S=0]=\beta_3+\beta_4+(\beta_6'+\beta_7')E[\bxy|S=1]
\end{equation*}
the result follows.

>>>>

\begin{prop}\label{prop:reg1}
  Under Assumptions \ref{ass:sm}, \ref{ass:rand}, \ref{ass:vps}, \ref{ass:ci}, and suitible regularity conditions \citep[e.g.][p. 327]{boosStefanskiBook}, if %principal scores are known exactly,
  $\ri$ is defined as in \eqref{eq:ri} and the model
\begin{equation}\label{eq:regression0}
  \EE[Y_i|Z_i,\ri]=\beta_0+\beta_1\ri+\beta_2Z_i+\beta_3Z_i\ri
\end{equation}
is fit with OLS, yielding coefficient estimates $\hat{\beta}_k$, $k=0,\dots,3$, then
\begin{equation}
  \begin{split}
    \hat{\eff0}&\equiv \hat{\beta}_2\\
    \hat{\eff1}&\equiv \hat{\beta}_3-\hat{\beta}_2
  \end{split}
\end{equation}
are M-estimators, with $\hat{\eff0}\rightarrow_p\eff0$ and $\hat{\eff1}\rightarrow_p\eff1$ as $n\rightarrow\infty$
\end{prop}

@@

The mixture estimator then fit two outcome models.
For the treatment group, a normal regression:
\begin{equation}\label{eq:mix1-sim}
  f_{Y|Z=1}(y_i)=\phi\left(\frac{y_i-[\beta_0+\beta_1S_i+\beta_2 +\bm{\gamma}_1^T\bm{x}_{i,1:2}]}{\sigma_T}\right)
\end{equation}
and for the control group, mixture model:
\begin{equation}\label{eq:mix-sim}
  f_{Y|Z=0}(y_i)=\ppf{\bm{\alpha}}{\bm{x}_{i,1:2}}\phi\left(\frac{y_i-[\beta_0+\beta_1+\bm{\gamma}_1^T\bm{x}_{i,1:2}]}{\sigma_C}\right)+(1-\ppf{\bm{\alpha}}{\bm{x}_{i,1:2}})\phi\left(\frac{y_i-[\beta_0+\bm{\gamma}_1^T\bm{x}_{i,1:2}]}{\sigma_C}\right)
\end{equation}
Note that the coefficients $\bm{\gamma}_1$ are the same across treatment groups and principal strata, while the residual variance differs across treatment groups, but not across principal strata.
The models \eqref{eq:psModsim}, \eqref{eq:mix1-sim}, and \eqref{eq:mix-sim} were fit simultaneously using Bayesian Markov Chain Monte Carlo with Stan \citep{rstan}. We interpreted the posterior mean as a point estimate and the posterior standard deviation as a standard error.

@@

There is good reason to hope that extensions to GEEPERs may be straightforward.
First, GEEPERs may be extended to situations in which there is two-way noncompliance, and only a weaker form of monotonicity (Assumption \ref{ass:sm}) holds.
For instance, in truncation by death problems \citep[e.g.][]{zhangRubin,ding2011}, $S_i=1$ if participant $i$ survives (or, more generally, if the outcome $Y_i$ is measured) and 0 otherwise, and interest is is in the principal effect for the ``always survive'' principal stratum in which $\sti=S_{ci}=1$.
If, say, $\sti\ge S_{Ci}$, then every subject in the control condition with $S_i=1$ is in the always survive stratum, while those subjects in the treatment condition with $S_i=1$ are a mixture of the always survive and $\sti=1;\;S_{Ci}=0$ stratum.
The scenario of truncation by death with (weak) monotonicity is broadly similar to the strong monotonicity, one-way noncompliance situation that this paper discussed.
On the other hand, when no monotonicity assumption holds, both $\st$ and $S_C$ will have to be imputed for every subject; further research is necessary to determine the appropriate way to do so.

@@

\subsection{Proof for Proposition \ref{prop:reg1}}
replacing $\sti$ and $\pp$ in \eqref{eq:estEq0} with $\ri$, as in \eqref{eq:ri}, and replacing $\tilde{\Psi}_i$ with $\Psi_i=\begin{psmallmatrix} 1 & 0&1&0\\ 0&1&0&1\\0&0&1&0\\0&0&0&1\end{psmallmatrix}\tilde{\Psi}_i$ gives an equivalent set of estimating equations:
\begin{equation}\label{eq:estEq1}
\sum_{i=1}^{n_C}\Psi_i=\sum_{i=1}^{n_C}\begin{pmatrix}
    Y_i-\muc0-\ri(\muc1-\muc0)-Z_i(\mut0-\muc0)-Z_i\ri(\mut1-\mut0-\muc1+\muc0)\\
    \ri Y_i-\ri\muc0-\ri^2(\muc1-\muc0)-Z_i\ri(\mut0-\muc0)-Z_i\ri^2(\mut0-\mut1-\muc0+\muc1)\\
    Z_iY_i-Z_i\mut0-Z_i\ri (\mut1-\mut0)\\
    Z_i\ri Y_i -Z_i\ri\mut0-Z_i\ri^2(\mut1-\mut0)

\end{pmatrix}=\begin{pmatrix} 0\\0\\0\\0\end{pmatrix}
\end{equation}
These are equivalent to the estimating equations for OLS model \eqref{eq:regression0} with $\beta_0=\muc0$, $\beta_1=\muc1-\muc0$, $\beta_2=\mut0-\muc0$, and $\beta_3=\mut1-\mut0-\muc1-\muc0$.
Therefore, under standard OLS regularity conditions the estimated parameter vector $\bm{\hat{\beta}}$ is consistent, completing the proof.

\subsection{Proof for Proposition \ref{prop:interactions}}
First, note that $\EE[S]=\EE\EE[S|\bx]=\EE[\pp]$ and
\begin{align*}
  \EE[\bx S]&=\EE[\bx S|Z=1]=\EE[\bx S|Z=0] \mbox{ due to \eqref{ass:rand}}\\
  &=\EE[\bx\EE[S|\bx]|Z=0]=\EE[\bx\pp|Z=0]=\EE[\bx\pp]
\end{align*}

implying that
\begin{equation*}
  \begin{split}
    \EE[Y|\bx,Z=0]&=\beta_0+\bm{\beta_1}\bxy+\beta_2\pp+\bm{\beta_5}'\bxy\pp\\
    \EE[Y|\bx,S,Z=1]&=\beta_0+\beta_3+(\bm{\beta_1}+\bm{\beta_6})\bxy+(\beta_3+\beta_4)S+(\bm{\beta_5}+\bm{\beta_7})'\bxy S
  \end{split}
\end{equation*}
Therefore,
\begin{align*}
  \EE[Y]&=\EE[Y|Z=0]+\EE[Z]\left\{\EE[Y|Z=1]-\EE[Y|Z=0]\right\}\\
  &=\beta_0+\bm{\beta_1}\EE[\bxy]+\beta_2\EE[\pp]+\bm{\beta_5}'\EE[\bxy\pp]+\EE[Z]\left\{\beta_3+\bm{\beta_4}\EE[S]+\bm{\beta_6}'\EE[\bxy]+\bm{\beta_7}'\EE[\bxy S]\right\}\\
  &=\beta_0+\bm{\beta_1}\EE[\bxy]+\beta_2\EE[R]+\bm{\beta_5}'\EE[\bxy R]+\EE[Z]\left\{\beta_3+\bm{\beta_4}\EE[R]+\bm{\beta_6}'\EE[\bxy]+\bm{\beta_7}'\EE[\bxy R]\right\}\\
\end{align*}
Analogous reasoning leads to expressions for $\EE[RY]$, $\EE[\bxy Y]$, $\EE[ZY]$,  $\EE[ZRY]$, $\EE[\bxy YZ]$, and $\EE[\bxy RZY]$.
These, in turn, give rise to estimating equations
\begin{equation*}
  \displaystyle\sum_i \begin{pmatrix}
    Y_i-\beta_0-\bm{\beta_1}\bxy_i-\beta_2R_i-\bm{\beta_5}'\bxy_iR_i-Z_i\left\{\beta_3-\bm{\beta_4}R_i-\bm{\beta_6}'\bxy_i-\bm{\beta_7}'\bxy_i R_i\right\}\\
    R_iY_i-\beta_0R_i-\bm{\beta_1}\bxy_i R_i-\beta_2R_i^2-\bm{\beta_5}'\bxy_iR_i^2-Z_i\left\{\beta_3R_i-\bm{\beta_4}R_i^2-\bm{\beta_6}'\bxy_iR_i-\bm{\beta_7}'\bxy_i R_i^2\right\}\\
    \bxy_i Y_i-\beta_0\bxy_i-\bm{\beta_1}\bxy_i\bx^{Y'}_i-\beta_2R_i\bxy_i-\bm{\beta_5}'\bxy_iR_i-Z_i\left\{\beta_3\bxy_i-\bm{\beta_4}R_i\bxy_i-\bm{\beta_6}'\bxy_i\bx^{Y'}_i-\bm{\beta_7}'\bxy_i\bx^{Y'}_i R_i\right\}\\
    Z_iY_i-Z_i\left\{\beta_0-\bm{\beta_1}\bxy_i-\beta_2R_i-\bm{\beta_5}'\bxy_iR_i-\beta_3-\bm{\beta_4}R_i-\bm{\beta_6}'\bxy_i-\bm{\beta_7}'\bxy_i R_i\right\}\\
    Z_iR_iY_i-Z_iR_i\left\{\beta_0-\bm{\beta_1}\bxy_i-\beta_2R_i-\bm{\beta_5}'\bxy_iR_i-\beta_3-\bm{\beta_4}R_i-\bm{\beta_6}'\bxy_i-\bm{\beta_7}'\bxy_i R_i\right\}\\
    Z_iY_i\bxy_i-Z_i\left\{\beta_0\bxy_i-\bm{\beta_1}\bxy_i\bx^{Y'}_i-\beta_2R_i\bxy_i-\bm{\beta_5}'\bxy_i\bx^{Y'}_iR_i-\beta_3\bxy_i-\bm{\beta_4}R_i\bxy_i-\bm{\beta_6}'\bxy_i\bx^{Y'}_i-\bm{\beta_7}'\bxy_i\bx^{Y'}_i R_i\right\}\\
    R_iZ_iY_i\bxy_i-Z_iR_i\left\{\beta_0\bxy_i-\bm{\beta_1}\bxy_i\bx^{Y'}_i-\beta_2R_i\bxy_i-\bm{\beta_5}'\bxy_i\bx^{Y'}_iR_i-\beta_3\bxy_i-\bm{\beta_4}R_i\bxy_i-\bm{\beta_6}'\bxy_i\bx^{Y'}_i-\bm{\beta_7}'\bxy_i\bx^{Y'}_i R_i\right\}
  \end{pmatrix}=\begin{pmatrix} 0\\0\\0\\0\\0\\0\\0\end{pmatrix}
\end{equation*}
These are the estimating equations for the regression model \eqref{eq:fullInteractionModel}, fit by OLS. Now note that
\begin{equation*}
  \EE[Y_T-Y_C|\st=0]=\beta_0^T-\beta_0^C+(\bm{\beta_1}^{T'}-\bm{\beta_1}^{C'})\EE[\bxy|S=0]=\beta_3+\beta_6'E[\bxy|S=0]
\end{equation*}
and
\begin{equation*}
  \EE[Y_T-Y_C|\st=1]=\beta_0^T+\beta_2^T-(\beta_0^C+\beta_2^C)+(\bm{\beta_1}^{T'}+\bm{\beta_3^{Ti}}-\bm{\beta_1}^{C'}-\bm{\beta_3^{C'}})\EE[\bxy|S=0]=\beta_3+\beta_4+(\beta_6'+\beta_7')E[\bxy|S=1]
\end{equation*}
the result follows.
